---
title: "Simulate"
author: "Leonardo Capitani"
date: "09/01/2025"
date-modified: last-modified
execute:
  echo: true
  warning: false
  message: false
format: 
      html:
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
      pdf:
             keep-tex: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load libraries

```{r load_packages}
library(here)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
library(GPEDM)
set.seed(123)
```

GGplot theme

```{r ggplot_theme}
newtheme <- theme_bw(base_size = 14) 
theme_set(newtheme)
```

## Simulate abiotic drivers

```{r}
# ---- 1. Simulate drivers ----
n <- 200
Time <- 1:n
# we assume no causal link from Temp to Nutrients 
temperature <- sin(2*pi*Time/50) + rnorm(n, 0, 0.3)   # seasonal + noise
phosphorus  <- sin(2*pi*Time/70 + 1) + rnorm(n, 0, 0.4)
nitrogen    <- sin(2*pi*Time/90 + 2) + rnorm(n, 0, 0.4)
```

```{r}
# ---- 2. Simulate emergent properties (hypotheses) ----
connectance <- 0.6*temperature + 0.3*nitrogen - 0.2*phosphorus +
               0.2*temperature*nitrogen + rnorm(n,0,0.2)

modularity  <- 0.5*phosphorus - 0.3*temperature + 
               0.1*nitrogen*phosphorus + rnorm(n,0,0.2)

stability   <- -0.6*temperature - 0.5*phosphorus - 0.4*nitrogen +
               0.3*temperature*phosphorus + rnorm(n,0,0.2)

df <- data.frame(Time, temperature, phosphorus, nitrogen,
                 connectance, modularity, stability)

```

Add quagga mussel abundance

```{r}

# ---- Quagga mussel abundance ----
quagga <- rep(NA,n) # NA observation before time 100
quagga[Time > 100] <- runif(sum(Time > 100), 20, 200)  # random after 100

# Standardize (z-score)
quagga_std <- scale(quagga)

df <- df %>% mutate(quagga = as.numeric(quagga_std))
# ---- Reshape emergent properties for plotting ----
df_long <- df %>%
  pivot_longer(cols = c(connectance, modularity, stability),
               names_to = "property", values_to = "value")

```

### Visualize data 

```{r plot_emergent_properties}


# ---- Plot with facet wrap ----
p <- ggplot(df_long, aes(x = Time)) +
  geom_line(aes(y = value, color = property), size = 1) +
  geom_col(data = df, aes(y = quagga, x = Time), 
           fill = "grey40", alpha = 0.4, inherit.aes = FALSE) +
  facet_wrap(~property, scales = "free_y", ncol = 1) +
  scale_y_continuous(
    name = "Plankton network emergent property (standardized)",
    sec.axis = sec_axis(~., name = "Quagga abundance (standardized)")
  ) +
  theme(legend.position = "none")

p
# save as PNG
ggsave(here("results","figures","simulations","emergent_properties.png"), plot = p, width = 8, height = 6, dpi = 300)
```

## Load lake Zurich data

```{r load_p}
p <- read.csv( here("Data","Merz","lake_zurich.csv")) |>   mutate(X = row_number())

# change species names
names(p)[names(p) == "Ci"] <- "Ciliates"
names(p)[names(p) == "Cr1"] <- "Cryptophytes (small)"
names(p)[names(p) == "Cr2"] <- "Cryptophytes (large)"
names(p)[names(p) == "Cy"] <- "Cyanobacteria"
names(p)[names(p) == "Di1"] <- "Diatoms (small)"
names(p)[names(p) == "Di2"] <- "Diatoms (large)"
names(p)[names(p) == "Go1"] <- "Gold algae (small)"
names(p)[names(p) == "Go2"] <- "Gold algae (large)"
names(p)[names(p) == "Gr1"] <- "Green algae (small)"
names(p)[names(p) == "Gr2"] <- "Green algae (large)"
names(p)[names(p) == "He"] <- "Large herbivores"
names(p)[names(p) == "Mi"] <- "Mixotrophic flagellates"
names(p)[names(p) == "Om"] <- "Omnivores"
names(p)[names(p) == "Pr"] <- "Invertebrate predators"
names(p)[names(p) == "Ro"] <- "Herbivore rotifers"
names(p)[names(p) == "temperature"] <- "Temperature"
names(p)[names(p) == "phosphorus"] <- "Phosphate"


```

### Visualize data

```{r plot_zurich_p, fig.width=12, fig.height=8}

# Separate abiotic and biotic variables
abiotic_vars <- c("Temperature", "Phosphate")

biotic_vars <- setdiff(names(p), c("X", "year", "month", "month_year", "quarter_year", abiotic_vars))

# Helper function to prepare p with stats
prepare_p <- function(df, vars) {
  df_long <- df %>%
    select(X, year, all_of(vars)) %>%
    pivot_longer(-c(X, year), names_to = "variable", values_to = "value")
  
  stats <- df_long %>%
    group_by(variable) %>%
    summarise(
      n_NA = sum(is.na(value)),
      n_obs = sum(!is.na(value)),
      start_year = min(year[!is.na(value)]),
      end_year = max(year[!is.na(value)]),
      .groups = "drop"
    )
  
  # Merge stats and calculate a fixed y-position (90% of max value)
  df_long <- df_long %>%
    left_join(stats, by = "variable") %>%
    group_by(variable) %>%
    mutate(y_pos = max(value, na.rm = TRUE) * 0.9) %>%
    ungroup()
  
  return(df_long)
}

# Prepare biotic and abiotic p
biotic_p <- prepare_p(p, biotic_vars)
abiotic_p <- prepare_p(p, abiotic_vars)

# Plot biotic variables
p_biotic <- ggplot(biotic_p, aes(x = X, y = value)) +
  geom_line(color = "steelblue") +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(
    aes(
      x = max(X, na.rm = TRUE),
      y = y_pos,
      label = paste0("NAs: ", n_NA, "\nObs: ", n_obs, "\nYears: ", start_year, "-", end_year)
    ),
    hjust = 1, vjust = 1, size = 3, inherit.aes = FALSE
  ) +
  labs(x = "Time", y = "Value")

#
# Print plots
p_biotic

```

```{r plot_abiotic}

p_abiotic <- ggplot(abiotic_p, aes(x = X, y = value)) +
  geom_line(color = "darkorange") +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(
    aes(
      x = max(X, na.rm = TRUE),
      y = y_pos,
      label = paste0("NAs: ", n_NA, "\nObs: ", n_obs, "\nYears: ", start_year, "-", end_year)
    ),
    hjust = 1, vjust = 1, size = 3, inherit.aes = FALSE
  ) +
  theme_minimal() +
  labs(x = "Time", y = "Value")

p_abiotic
```

## Model plankton network

## Model emergent properties

IfÂ `E`Â andÂ `tau`Â are supplied, the model uses the E/tau Method. If using the E/tau Method and multiple variables are listed underÂ `x`,Â `fitGP`Â usesÂ ð¸Â lags with spacingÂ ðœÂ ofÂ *each*Â variable. IfÂ `E`Â andÂ `tau`Â are omitted, the function will use the Custom Lags Method, using all the variables inÂ `x`Â as predictors.

```{r fit_GPEDM}

m0 <- fitGP(data = df, 
            
            
                  time = "Time",
                   
                  y = "connectance", x = c("nitrogen", "phosphorus", "temperature"),
                   
                   E = 1, tau = 1,
                   
                  predictmethod = "loo", exclradius = 6)


```

## Conditional effects

```{r}

con = getconditionals(m0)

```

Predictions:

```{r}
plot(m0)
```
