---
title: "Registered Report notebook: simulate"
author: 
- name: Leonardo Capitani 
  email: leonardo.capitani@eawag.ch
date: "09/01/2025"
date-modified: last-modified
execute:
  echo: true
  warning: false
  message: false
format: 
      html:
          css: styles.css
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
      pdf:
             keep-tex: true
editor: visual
editor_options: 
  chunk_output_type: console
output-dir: docs
bibliography: references.bib
csl: eco_lett.csl
---

## Intro

I made this notebook to visualize how we are going to answer the research questions of the research project: Inferring and predicting plankton networks emergent properties in changing lakes.

::: {.callout-note appearance="minimal"}
**Research questions:**

1.  How do temperature, phosphorus and nitrogen levels have caused network emergent properties to vary?

2.  How do invasive species presence/abundance have influenced network emergent properties?

3.  How will plankton network emergent properties change in the near future (next 5-10 years)?
:::

::: {.callout-important appearance="minimal"}
**Plankton emergent properties**

-   Connectance (C): percentage of associations between network nodes. Defined by [@merz2023] as: $C = 100 \times \frac{L}{N(N - 1)}$, where $L$ is the number of interactions between nodes and $N$ is the number of nodes in the system;

-   Modularity (M): the strength of division of a network into modules. Modularity is a network property that describes the tendency of nodes to cluster. Networks with high modularity have dense connections among nodes within modules but sparse connections to nodes in other modules [@gilarranz2017]. In order to calculate network modularity, we have to find a network partition into modules that maximize the value of modularity. A partition is just a method of assigning nodes into modules. This partition is the one that groups the nodes in a way in which the fraction of links between modules is minimal when compared with the fraction of links inside modules. We define the partition as [@guimerà2005] did:

    $$
    M \equiv \sum_{s=1}^{N_M} 
    \left[ 
    \frac{l_s}{L} - \left( \frac{d_s}{2L} \right)^2 
    \right]
    \tag{1}
    $$

    Where:

    -   $N_M$ is the number of modules\
    -   $L$ is the number of links in the entire component\
    -   $l_s$ is the number of links between nodes that belong to module $s$\
    -   $d_s$ is the sum of the degrees of the nodes that belong to module $s$

-   Stability: we define the network stability as the volume contraction rate (VCR) of a non-equilibrium dynamical system [@cenci2019]. We will measure the VCR using the trace of the Jacobian species interaction matrix (sum of the first diagonal). The network stability metric we are proposing combines and summarises species interdependencies (biotic interactions) and single-species sensitivities to environmental change (intrinsic growth). The smaller the VCR, the larger the network stability.
:::

## Load libraries

```{r load_packages}
library(here)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
library(GPEDM)
set.seed(123)
```

GGplot theme

```{r ggplot_theme}
newtheme <- theme_bw(base_size = 14) 
theme_set(newtheme)
```

## Simulate abiotic drivers

```{r}
# ---- 1. Simulate drivers ----
n <- 200
Time <- 1:n
# we assume no causal link from Temp to Nutrients 
temperature <- sin(2*pi*Time/50) + rnorm(n, 0, 0.3)   # seasonal + noise
phosphorus  <- sin(2*pi*Time/70 + 1) + rnorm(n, 0, 0.4)
nitrogen    <- sin(2*pi*Time/90 + 2) + rnorm(n, 0, 0.4)
```

```{r}
# ---- 2. Simulate emergent properties (hypotheses) ----
connectance <- 0.6*temperature + 0.3*nitrogen - 0.2*phosphorus +
               0.2*temperature*nitrogen + rnorm(n,0,0.2)

modularity  <- 0.5*phosphorus - 0.3*temperature + 
               0.1*nitrogen*phosphorus + rnorm(n,0,0.2)

stability   <- -0.6*temperature - 0.5*phosphorus - 0.4*nitrogen +
               0.3*temperature*phosphorus + rnorm(n,0,0.2)

df <- data.frame(Time, temperature, phosphorus, nitrogen,
                 connectance, modularity, stability)

```

Add quagga mussel abundance

```{r}

# ---- Quagga mussel abundance ----
quagga <- rep(NA,n) # NA observation before time 100
quagga[Time > 100] <- runif(sum(Time > 100), 20, 200)  # random after 100

# Standardize (z-score)
quagga_std <- scale(quagga)

df <- df %>% mutate(quagga = as.numeric(quagga_std))
# ---- Reshape emergent properties for plotting ----
df_long <- df %>%
  pivot_longer(cols = c(connectance, modularity, stability),
               names_to = "property", values_to = "value")

```

### Visualize data

#### Time vs emergent properties

```{r plot_emergent_properties}


# ---- Plot with facet wrap ----
p <- ggplot(df_long, aes(x = Time)) +
  geom_line(aes(y = value, color = property), size = 1) +
  geom_col(data = df, aes(y = quagga, x = Time), 
           fill = "grey40", alpha = 0.4, inherit.aes = FALSE) +
  facet_wrap(~property, scales = "free_y", ncol = 1) +
  scale_y_continuous(
    name = "Plankton network emergent property (standardized)",
    sec.axis = sec_axis(~., name = "Quagga abundance (standardized)")
  ) +
  theme(legend.position = "none")

p
# save as PNG
ggsave(here("outputs","figures","simulations","emergent_properties.png"), plot = p, width = 8, height = 6, dpi = 300)
```

#### Heat maps

## Load lake Zurich data

```{r load_p}
p <- read.csv( here("data","data_raw","zurich","lake_zurich.csv")) |>   mutate(X = row_number())

# change species names
names(p)[names(p) == "Ci"] <- "Ciliates"
names(p)[names(p) == "Cr1"] <- "Cryptophytes (small)"
names(p)[names(p) == "Cr2"] <- "Cryptophytes (large)"
names(p)[names(p) == "Cy"] <- "Cyanobacteria"
names(p)[names(p) == "Di1"] <- "Diatoms (small)"
names(p)[names(p) == "Di2"] <- "Diatoms (large)"
names(p)[names(p) == "Go1"] <- "Gold algae (small)"
names(p)[names(p) == "Go2"] <- "Gold algae (large)"
names(p)[names(p) == "Gr1"] <- "Green algae (small)"
names(p)[names(p) == "Gr2"] <- "Green algae (large)"
names(p)[names(p) == "He"] <- "Large herbivores"
names(p)[names(p) == "Mi"] <- "Mixotrophic flagellates"
names(p)[names(p) == "Om"] <- "Omnivores"
names(p)[names(p) == "Pr"] <- "Invertebrate predators"
names(p)[names(p) == "Ro"] <- "Herbivore rotifers"
names(p)[names(p) == "temperature"] <- "Temperature"
names(p)[names(p) == "phosphorus"] <- "Phosphate"


```

### Visualize data

```{r plot_zurich_p, fig.width=12, fig.height=8}

# Separate abiotic and biotic variables
abiotic_vars <- c("Temperature", "Phosphate")

biotic_vars <- setdiff(names(p), c("X", "year", "month", "month_year", "quarter_year", abiotic_vars))

# Helper function to prepare p with stats
prepare_p <- function(df, vars) {
  df_long <- df %>%
    select(X, year, all_of(vars)) %>%
    pivot_longer(-c(X, year), names_to = "variable", values_to = "value")
  
  stats <- df_long %>%
    group_by(variable) %>%
    summarise(
      n_NA = sum(is.na(value)),
      n_obs = sum(!is.na(value)),
      start_year = min(year[!is.na(value)]),
      end_year = max(year[!is.na(value)]),
      .groups = "drop"
    )
  
  # Merge stats and calculate a fixed y-position (90% of max value)
  df_long <- df_long %>%
    left_join(stats, by = "variable") %>%
    group_by(variable) %>%
    mutate(y_pos = max(value, na.rm = TRUE) * 0.9) %>%
    ungroup()
  
  return(df_long)
}

# Prepare biotic and abiotic p
biotic_p <- prepare_p(p, biotic_vars)
abiotic_p <- prepare_p(p, abiotic_vars)

# Plot biotic variables
p_biotic <- ggplot(biotic_p, aes(x = X, y = value)) +
  geom_line(color = "steelblue") +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(
    aes(
      x = max(X, na.rm = TRUE),
      y = y_pos,
      label = paste0("NAs: ", n_NA, "\nObs: ", n_obs, "\nYears: ", start_year, "-", end_year)
    ),
    hjust = 1, vjust = 1, size = 3, inherit.aes = FALSE
  ) +
  labs(x = "Time", y = "Value")

#
# Print plots
p_biotic

```

```{r plot_abiotic}

p_abiotic <- ggplot(abiotic_p, aes(x = X, y = value)) +
  geom_line(color = "darkorange") +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(
    aes(
      x = max(X, na.rm = TRUE),
      y = y_pos,
      label = paste0("NAs: ", n_NA, "\nObs: ", n_obs, "\nYears: ", start_year, "-", end_year)
    ),
    hjust = 1, vjust = 1, size = 3, inherit.aes = FALSE
  ) +
  theme_minimal() +
  labs(x = "Time", y = "Value")

p_abiotic
```

## Model plankton network

## Model emergent properties

If `E` and `tau` are supplied, the model uses the E/tau Method. If using the E/tau Method and multiple variables are listed under `x`, `fitGP` uses 𝐸 lags with spacing 𝜏 of *each* variable. If `E` and `tau` are omitted, the function will use the Custom Lags Method, using all the variables in `x` as predictors.

```{r fit_GPEDM}

m0 <- fitGP(data = df, 
            
            
                  time = "Time",
                   
                  y = "connectance", x = c("nitrogen", "phosphorus", "temperature"),
                   
                   E = 1, tau = 1,
                   
                  predictmethod = "loo", exclradius = 6)


```

## Conditional effects

```{r}

con = getconditionals(m0)

```

Predictions:

```{r}
plot(m0)
```

## References
